Copyright

### 2025-02-08 code.qmd
Minor improvements
Reformat markdown
To ease change tracking in the future.
Too bad it wasn't done from the
beginning.

- Data may be collected from social network sites using their APIs^[An **a**pplication **p**rogramming **i**nterface provides the capability to interact with other software using a programming language.] or downloaded from a data repository, or
- an analysis may be conducted with the help of AI providers.

@@

To format manuscripts according to the requirements of a particular journal, [multiple Quarto extensions are available](https://quarto.org/docs/extensions/listing-journals.html). In the following, we will demonstrate the usage of [`apaquarto`](https://wjschne.github.io/apaquarto/), which typesets documents according to the requirements of the American Psychological Association [-@APA2020]. Of course, you may also use an extension for a different journal.

@@

@@

> **1\. Importance**: Software should be considered a legitimate and citable product of research. Software citations should be accorded the same importance in the scholarly record as citations of other research products, such as publications and data; they should be included in the metadata of the citing work, for example in the reference list of a journal article, and should not be omitted or separated. Software should be cited on the same basis as any other research product such as a paper or a book, that is, authors should cite the appropriate set of software products just as they cite the appropriate set of papers.
> 
> **5\. Accessibility**: Software citations should facilitate access to the software itself and to its associated metadata, documentation, data, and other materials necessary for both humans and machines to make informed use of the referenced software.
> 
> **6\. Specificity**: Software citations should facilitate identification of, and access to, the specific version of software that was used. Software identification should be as specific as necessary, such as using version numbers, revision numbers, or variants such as platforms.

In practice, you would identify all pieces of software the project relies on. A few of them are obvious, such as R itself, Quarto, and the $\TeX$ distribution we installed before. Then there are the individual R packages^[By default, `renv` avoids copying them into the project folder. And even that happens (e.g., by using `renv::isolate()`) they are excluded from Git by default.], Quarto extensions, and $\TeX$ packages. All of them, in turn, may have dependencies and it is up to you decide when not to dig deeper. For example, some R packages are only thin wrappers around other R packages or around system dependencies which also might deserve credit. A system dependency is additional software that you require on your computer apart from the R package.

@@

"Should I cite the software?" by @Brown2016 licensed under [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/). Simplified from original.
:::

Now, add references for the software you would like to cite to the manuscript. In the following, we will demonstrate this for R and all R packages by using the R package `grateful`. For arbitrary software, you can use [CiteAs](https://citeas.org/) or the [DOI Citation Formatter](https://citation.doi.org/) to create appropriate citations.

@@

This will automatically create a paragraph citing all used packages and generate the bibliography file `grateful-refs.bib`.^[Note that this automatic detection can miss packages in some circumstances, therefore always verify the rendered result.] Then, in the YAML header, add `grateful-refs.bib` by setting the `bibliography` as follows:

@@

- __Don't reinvent the wheel.__ With R, chances are that what you need to do is greatly facilitated by a package from one of many high-quality collections such as [rOpenSci](https://ropensci.org/packages/), [r-lib](https://github.com/r-lib), [Tidyverse](https://www.tidyverse.org/packages/), or [fastverse](https://fastverse.github.io/fastverse/).

- __Think twice about your dependencies.__ Every dependency increases the risk of irreproducibility in the future. Prefer packages that are well-maintained and light on dependencies^[You can use the function `pak::pkg_deps()` to count the total number of package dependencies in R.]. We also recommend you to read "When should you take a dependency?" by @Wickham2023PackagesDependency.

@@

- __Test your code.__ Test your code with scenarios where you know what the result should be. Turn bugs you discovered into test cases. Use linting tools^[A linting tool analyzes your code without actually running it. Therefore, this process is also called static code analysis.] to identify common mistakes in your code, for example, the R package [`lintr`](https://lintr.r-lib.org/).

@@

@@

> "Any fool can write code that a computer can understand. Good programmers write code that humans can understand."
> 
> --- @Fowler1999, p. 15

@@

This will create a few additional files which facilitate reconstructing the computational environment in the future.^[Either using [repo2docker](https://repo2docker.readthedocs.io/) or the public [binder](https://mybinder.org/) service.] As always, commit your changes:

>>>>

- Data may be collected from social network sites using their APIs[^api-explanation]
  or downloaded from a data repository, or
- an analysis may be conducted with the help of AI providers.

[^api-explanation]: An **a**pplication **p**rogramming **i**nterface
provides the capability to interact with other software using a programming language.

@@

To format manuscripts according to the requirements of a particular journal,
[multiple Quarto extensions are available][apaquarto-extensions].
In the following, we will demonstrate the usage of [`apaquarto`][apaquarto],
which typesets documents according to the requirements
of the American Psychological Association [-@APA2020].
Of course, you may also use an extension for a different journal.

[apaquarto]: https://wjschne.github.io/apaquarto/
[apaquarto-extensions]: https://quarto.org/docs/extensions/listing-journals.html

@@

It is best to provide attribution even if the license does not require it.

@@

> **1\. Importance**: Software should be considered a legitimate and citable
> product of research. Software citations should be accorded the same importance
> in the scholarly record as citations of other research products, such as
> publications and data; they should be included in the metadata of the citing
> work, for example in the reference list of a journal article, and should not
> be omitted or separated. Software should be cited on the same basis as any
> other research product such as a paper or a book, that is, authors should cite
> the appropriate set of software products just as they cite the appropriate set
> of papers.
> 
> **5\. Accessibility**: Software citations should facilitate access to the
> software itself and to its associated metadata, documentation, data, and other
> materials necessary for both humans and machines to make informed use of the
> referenced software.
> 
> **6\. Specificity**: Software citations should facilitate identification of,
> and access to, the specific version of software that was used. Software
> identification should be as specific as necessary, such as using version
> numbers, revision numbers, or variants such as platforms.

In practice, you would identify all pieces of software the project relies on.
A few of them are obvious, such as R itself, Quarto,
and the $\TeX$ distribution we installed before.
Then there are the individual R packages[^renv-copying],
Quarto extensions, and $\TeX$ packages.
All of them, in turn, may have dependencies
and it is up to you decide when not to dig deeper.
For example, some R packages are only thin wrappers around other R packages
or around system dependencies which also might deserve credit.
A system dependency is additional software
that you require on your computer apart from the R package.

[^renv-copying]: By default, `renv` avoids copying them into the project folder.
And even that happens (e.g., by using `renv::isolate()`)
they are excluded from Git by default.

@@

"Should I cite the software?" by @Brown2016 licensed under [CC\ BY-SA\ 4.0][cc-by-sa].
Simplified from original.
:::

[cc-by-sa]: https://creativecommons.org/licenses/by-sa/4.0/

Now, add references for the software you would like to cite to the manuscript.
In the following, we will demonstrate this for R and all R packages
by using the R package `grateful`.
For arbitrary software,
you can use [CiteAs][citeas] or the [DOI Citation Formatter][doi-citation]
to create appropriate citations.

[citeas]: https://citeas.org/
[doi-citation]: https://citation.doi.org/

@@

This will automatically create a paragraph citing all used packages
and generate the bibliography file `grateful-refs.bib`.[^automatic.note]
Then, in the YAML header,
add `grateful-refs.bib` by setting the `bibliography` as follows:

[^automatic.note]: Note that this automatic detection
can miss packages in some circumstances,
therefore always verify the rendered result.

@@

- __Don't reinvent the wheel.__
  With R, chances are that what you need to do is greatly facilitated
  by a package from one of many high-quality collections
  such as [rOpenSci][ropensci], [r-lib][r-lib], [Tidyverse][tidyverse],
  or [fastverse][fastverse].

- __Think twice about your dependencies.__
  Every dependency increases the risk of irreproducibility in the future.
  Prefer packages that are well-maintained and light on dependencies[^pkg-deps].
  We also recommend you to read "When should you take a dependency?"
  by @Wickham2023PackagesDependency.

@@

- __Test your code.__
  Test your code with scenarios where you know what the result should be.
  Turn bugs you discovered into test cases. Use linting tools[^linting-explanation]
  to identify common mistakes in your code,
  for example, the R package [`lintr`](https://lintr.r-lib.org/).

@@

[ropensci]: https://ropensci.org/packages/
[r-lib]: https://github.com/r-lib
[tidyverse]: https://www.tidyverse.org/packages/
[fastverse]: https://fastverse.github.io/fastverse/

[^pkg-deps]: You can use the function `pak::pkg_deps()`
to count the total number of package dependencies in R.

[^linting-explanation]: A linting tool analyzes your code without actually running it. Therefore, this process is also called static code analysis.

@@

> "Any fool can write code that a computer can understand.
> Good programmers write code that humans can understand."
> 
> --- @Fowler1999, p. 15

@@

This will create a few additional files which facilitate
reconstructing the computational environment in the future.[^binder-note]
As always, commit your changes:

[^binder-note]: Either using [repo2docker](https://repo2docker.readthedocs.io/)
or the public [binder](https://mybinder.org/) service.

### 2025-02-08 copyright.qmd
Reformat markdown
To ease change tracking in the future.
Too bad it wasn't done from the
beginning.

The following video "The correct indication of a license" embedded from the TIB AV-Portal ([Data Privacy Statement](https://av.tib.eu/privacy)) explains the correct attribution of Creative Commons licenses.

@@

Note that copyright law [varies between countries](https://commons.wikimedia.org/wiki/Commons:Copyright_rules_by_territory) and the rule of territoriality (_lex loci protectionis_) applies: Applicable are the laws of the country where a work is used. Also, if the work was created as part of the author's job, it might be their employer who holds the copyright to the work, depending on the country and contract.

A license is a legal document that regulates what other people than the author can and cannot do with a copyright-protected work, and under which conditions. The license may be exclusive, such that it is only granted to one recipient, effectively limiting the author in dealing with their own work. It can also be non-exclusive, thus giving rights to others without the author losing them. To avoid that every author who would like to share their work writes their own license, pre-formulated "boilerplate" licenses have been created. Recognizing them facilitates both sharing (on the side of the author) and re-using (on the side of the recipient).^[In fact, authors should not create their own license texts if the purpose is to share their work freely and openly with others. By using a boilerplate license others know what to expect, also because it may have already been [tested](https://sfconservancy.org/copyleft-compliance/) in [courts](https://legaldb.creativecommons.org/).]

@@

The terms _free_ and _open_, especially with regard to software, come with a history. The Free Software Foundation (FSF) was founded in 1985 to protect "four essential freedoms" [-@FSF2024] of a program's user. These are the freedoms to use, study, share, and improve a program. Software whose users legally and practically have these freedoms (because, among other things, they have access to its source code) is considered _free_. The four freedoms are seen as vital for a society as a whole in the sense that they enable sharing, cooperation and ultimately freedom in general. The FSF maintains [a list of software licenses](https://www.gnu.org/licenses/license-list.en.html#SoftwareLicenses) that it considers to be protecting the four freedoms. Sometimes the term _libre_ (Spanish and French for _free_) is used to make a distinction from _gratis_ software. You can learn more about free software at [Write Free Software](https://writefreesoftware.org/learn).

The Open Source Initiative (OSI), which was founded in 1998, follows a more pragmatic approach. It is concerned with developing high-quality software, for which everyone's ability to obtain, modify and contribute back the source code is considered beneficial. Access to the source code is one out of multiple conditions for software to be considered _open source_ by the OSI [-@OSI2007], which equally maintains [a list of approved licenses](https://opensource.org/licenses).

@@

To determine whether a particular work can be included in the project folder and shared publicly, the first step is to search for its license. For example, this tutorial has an [about page](about.qmd) detailing its licensing, whereas other websites sometimes state the license in the footer. Other places to look include a project's `README` file or a dedicated `LICENSE` or `COPYING` file. If a project has been stored in a research repository, yet another means of communicating its copyright status might exist. For media from Wikimedia Commons, an [Attribution Generator](https://lizenzhinweisgenerator.de/?lang=en) exists.

@@

If you cannot find any license, the work might be in the public domain^[in Germany: _Gemeinfreiheit_] depending on national law. For example, the work may not cross the [threshold of originality](https://commons.wikimedia.org/wiki/Commons:Threshold_of_originality), it may not [qualify for copyright protection](https://commons.wikimedia.org/wiki/Commons:Unprotected_works) -- such as official works by the government --, or the copyright protection might have expired. According to the [Berne convention](https://en.wikipedia.org/wiki/Berne_Convention), copyright expires at earliest 50 years after the author's death, but in [many countries](https://commons.wikimedia.org/wiki/Help:Public_domain#Copyright_terms_by_country) it expires even later -- you can use a [public domain calculator](https://lawflow.org/public-domain-calculators/) to check. If no license notice is present and the work is not in the public domain, it is _in copyright_ and you are generally not allowed to share it with others. You can, however, ask the author to put the corresponding work under a free/open license -- Wikimedia Commons provides an [email template](https://commons.wikimedia.org/wiki/Commons:Email_templates) to obtain such a permission. Without permission you can check you national laws for limitations and exceptions to copyright.^[in Germany: _Schranken des Urheberrechts_] Examples include the right to quote, [trivial use](https://commons.wikimedia.org/wiki/Commons:De_minimis), and the US-American [fair use](https://en.wikipedia.org/wiki/Wikipedia:FAQ/Copyright#What_is_fair_use?).

@@

For a software license to be considered free/open, it must be approved by either the [Open Source Initiative](https://opensource.org/licenses) (OSI) or the [Free Software Foundation](https://www.gnu.org/licenses/license-list.html#SoftwareLicenses) (FSF). Some of the most common free/open software licenses are listed below:

@@

  [Link to full text](https://creativecommons.org/publicdomain/zero/1.0/legalcode#legal-code-title)
  </details>
- <details>
  <summary>Creative Commons Attribution 4.0 (`CC-BY-4.0`)</summary>
  __Major conditions when sharing works__:

@@

  [Link to full text](https://creativecommons.org/licenses/by/4.0/legalcode#legal-code-title)
  </details>
- <details>
  <summary>Creative Commons Attribution-ShareAlike 4.0 (`CC-BY-SA-4.0`)</summary>
  __Major conditions when sharing works__:

@@

  [Link to full text](https://creativecommons.org/licenses/by-sa/4.0/legalcode#legal-code-title)
  </details>
::::

Of course, if you encounter a Creative Commons license with a previous version that satisfies the description above, it can be also considered free/open. If you encounter a license not from Creative Commons and want to determine whether it is free/open, you can consult the [lists of licenses accepted at Wikimedia Commons](https://commons.wikimedia.org/wiki/Commons:Copyright_tags). Note, however, that some of the licenses there might be inappropriate for text and media. For example, licenses for software or for documentation commonly require to distribute a work together with the full license text, which is impractical for usage in print media.

@@

  [Link to full text](https://creativecommons.org/licenses/by-sa/4.0/legalcode)
  </details>
::::

@@

If you encounter a license not listed here and want to determine whether it is free/open, you can consult the [lists of licenses accepted at Wikimedia Commons](https://commons.wikimedia.org/wiki/Commons:Copyright_tags). Note, however, that some of the licenses there might be inappropriate for data, because they do not mention _sui generis_ database rights.
:::

@@

You may have noticed that in many cases a version number is part of the license name. This is because the organizations that created the licenses sometimes publish updated versions to accommodate for developments in copyright law and the communities that use the licenses. For example, the Creative Commons licenses (that start with `CC`) were [first published in 2002](https://creativecommons.org/weblog/entry/3476). Since then, the possibility to relicense under later ([v2.0](https://creativecommons.org/weblog/entry/4216#:~:text=Share%20Alike%20Across%20Borders), 2004) and compatible licenses ([v3.0](https://creativecommons.org/weblog/entry/7249#:~:text=BY-SA%20%E2%80%94%20Compatibility%20Structure%20Now%20Included), 2007) has been added, a 30-day window to correct license violations has been established to combat [copyleft trolls](https://commons.wikimedia.org/wiki/Commons:Copyleft_trolling), and _sui generis_ database rights have been covered explicitly ([v4.0](https://creativecommons.org/weblog/entry/40768), 2013). There are many more [subtle differences between license versions](https://wiki.creativecommons.org/wiki/License_Versions), therefore it is important to indicate which license version exactly one is referring to, as the license of a work does not "update" automatically (i.e., you have to follow the terms of that exact version). CC0, officially introduced in 2009, is still at version [1.0](http://www.creativecommons.org/weblog/entry/13304).

@@

Finally, if you verified that the work has been made available to you under a free/open license, you need to understand the conditions under which you are allowed to include the work in your project folder and share it with others -- regardless of whether you modified it. In the following, we will explain the most important conditions of the licenses listed above. For other licenses we recommend reading their text in full and/or consulting with a lawyer. The conditions of many popular free/open licenses are also [summarized on ChooseALicense.com](https://choosealicense.com/appendix/), which you can use to get a general idea of a license (but which does not obviate reading the license text in full). Specifically for software licenses the _Open Source Automation Development Lab_ provides [useful checklists detailing the respective obligations](https://www.osadl.org/Checklists/).^[Accessing the checklists requires a (free) account on their website.]

@@

Licenses may require that you provide attribution to the author(s)^[or others whose names are provided with the original work] and to retain any copyright notices,^[Such as `Copyright (c) 2025 Jane Doe`] although the specifics vary between licenses. For example, Creative Commons licenses other than CC0 require that the name of the author, the title of the work, and a link to the work are provided.

@@

If you modified the work and put a sufficient amount of creativity into it, the result is called a _derivative work_. It is (again) protected by copyright and even if you share it, nobody except you is allowed to copy, modify, or distribute it further. You can put it under a free/open license but don't have to.^[Of course, you still need to abide by the license conditions of the original work, such as attribution.] The only exception is if the original work has been made available under a "copyleft" license. In that case, if you share the work with others you must put it under the same license as the original work^[or sometimes also under a later version or under a compatible license] -- or you commit a copyright violation.^[Note, however, that [copyleft licenses are not automatic](https://commons.wikimedia.org/wiki/Commons:Viral_licenses_are_not_automatic): A derivative work of a copylefted work is not "implicitly" under the same license -- others would commit a copyright violation if they just took it. Only if the derivative author explicitly makes it available under the same license, others may use the derivative work under the same terms.] This explains the origin of the term _copyleft_, as it "turns copyright around" to make works permanently free: ©\ →\ 🄯. In the realm of Creative Commons licenses, copyleft is also called _share alike_. The opposite of a copyleft license is sometimes also called a _permissive_ license.

What is considered creation of a derivative work varies between jurisdictions, but the following are typical examples: translating a text into a different language, taking a picture of an artwork, adding a song to a video, adapting a computer program to fit own use cases, or taking a [screenshot of a computer program](https://commons.wikimedia.org/wiki/Commons:Screenshots). In all of these cases, if the original work is under a copyleft license, then sharing of the resulting derivative work is only allowed if it is put under the same license.

@@

Obviously, if you are in doubt whether a condition of the license applies in your use case, it is better to either follow it or to ask the author for clarification or explicit permission. This is especially important given the existence of [_copyleft trolls_](https://commons.wikimedia.org/wiki/Commons:Copyleft_trolling): People who distribute works under a free/open license but send out statutory damages claims after minor violations without providing reusers an opportunity to fix their error. In addition, in many European countries attribution may be necessary regardless due to the author's moral rights. For these reasons, we recommend that you provide attribution for all works under a free/open license -- even if the license does not strictly require it (e.g., CC0). For this reason, you need to make note of every foreign work included in your project folder and record its license.

@@

The free/open licenses we discuss here mostly grant copyrights (with some of them explicitly granting patent rights and _sui generis_ database rights). Therefore, you may lack other rights such as personality, privacy, moral, or trademark rights. For example, [sharing photos that depict people](https://commons.wikimedia.org/wiki/Commons:Photographs_of_identifiable_people) is not only a matter of copyright, but also of privacy rights. Also, their commercial use may require the consent of the depicted person. Finally, note that even if attribution is not a requirement of the license, good scientific practice demands that appropriate citations are made.
:::

@@

1. The picture has been put under the license [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/)
2. The license is free/open
3. The license is appropriate for images
4. The license contains no (major) conditions
:::

>>>>

The following video "The correct indication of a license"
embedded from the TIB AV-Portal ([Data Privacy Statement](https://av.tib.eu/privacy))
explains the correct attribution of works under Creative Commons licenses.

@@

Note that copyright law [varies between countries][commons-rules]
and the rule of territoriality (_lex loci protectionis_) applies:
Applicable are the laws of the country where a work is used.
Also, if the work was created as part of the author's job,
it might be their employer who holds the copyright to the work,
depending on the country and contract.

[commons-rules]: https://commons.wikimedia.org/wiki/Commons:Copyright_rules_by_territory

A license is a legal document that regulates what other people than the author
can and cannot do with a copyright-protected work, and under which conditions.
The license may be exclusive, such that it is only granted to one recipient,
effectively limiting the author in dealing with their own work.
It can also be non-exclusive, thus giving rights to others without the author losing them.
To avoid that every author who would like to share their work writes their own license,
pre-formulated "boilerplate" licenses have been created.
Recognizing them facilitates both sharing (on the side of the author)
and re-using (on the side of the recipient).^[In fact, authors should not
create their own license texts if the purpose is to share their work freely
and openly with others.
By using a boilerplate license others know what to expect,
also because it may have already been [tested][sfconservancy] in [courts][cc-db].]

[sfconservancy]: https://sfconservancy.org/copyleft-compliance/
[cc-db]: https://legaldb.creativecommons.org/

@@

The terms _free_ and _open_, especially with regard to software, come with a history.
The Free Software Foundation (FSF) was founded in 1985
to protect "four essential freedoms" [-@FSF2024] of a program's user.
These are the freedoms to use, study, share, and improve a program.
Software whose users legally and practically have these freedoms
(because, among other things, they have access to its source code) is considered _free_.
The four freedoms are seen as vital for a society as a whole
in the sense that they enable sharing, cooperation and ultimately freedom in general.
The FSF maintains [a list of software licenses][fsf-list]
that it considers to be protecting the four freedoms.
Sometimes the term _libre_ (Spanish and French for _free_)
is used to make a distinction from _gratis_ software.
You can learn more about free software at [Write Free Software](https://writefreesoftware.org/learn).

[fsf-list]: https://www.gnu.org/licenses/license-list.html#SoftwareLicenses

The Open Source Initiative (OSI), which was founded in 1998, follows a more pragmatic approach.
It is concerned with developing high-quality software,
for which everyone's ability to obtain, modify and contribute back the source code
is considered beneficial.
Access to the source code is one out of multiple conditions for software
to be considered _open source_ by the OSI [-@OSI2007],
which equally maintains [a list of approved licenses][osi-list].

[osi-list]: https://opensource.org/licenses

@@

To determine whether a particular work can be included in the project folder
and shared publicly, the first step is to search for its license.
For example, this tutorial has an [about page](about.qmd) detailing its licensing,
whereas other websites sometimes state the license in the footer.
Other places to look include a project's `README` file or a dedicated `LICENSE` or `COPYING` file.
If a project has been stored in a research repository,
yet another means of communicating its copyright status might exist.
For media from Wikimedia Commons, an [Attribution Generator][commons-gen] exists.

[commons-gen]: https://lizenzhinweisgenerator.de/?lang=en

@@

If you cannot find any license, the work might be
in the public domain^[in Germany: _Gemeinfreiheit_] depending on national law.
For example, the work may not cross the [threshold of originality][commons-threshold],
it may not [qualify for copyright protection][commons-unprotected]
-- such as official works by the government --,
or the copyright protection might have expired.
According to the [Berne convention][berne-convention],
copyright expires at earliest 50 years after the author's death,
but in [many countries][commons-pd] it expires even later
-- you can use a [public domain calculator][pd-calc] to check.
If no license notice is present and the work is not in the public domain,
it is _in copyright_ and you are generally not allowed to share it with others.
You can, however, ask the author to put the corresponding work under a free/open license
-- Wikimedia Commons provides an [email template][commons-templates] to obtain such a permission.
Without permission you can check you national laws
for limitations and exceptions to copyright.^[in Germany: _Schranken des Urheberrechts_]
Examples include the right to quote, [trivial use][commons-minimis],
and the US-American [fair use][wiki-fair-use].

[commons-threshold]: https://commons.wikimedia.org/wiki/Commons:Threshold_of_originality
[commons-unprotected]: https://commons.wikimedia.org/wiki/Commons:Unprotected_works
[berne-convention]: https://en.wikipedia.org/wiki/Berne_Convention
[commons-pd]: https://commons.wikimedia.org/wiki/Help:Public_domain#Copyright_terms_by_country
[pd-calc]: https://lawflow.org/public-domain-calculators/
[commons-templates]: https://commons.wikimedia.org/wiki/Commons:Email_templates
[commons-minimis]: https://commons.wikimedia.org/wiki/Commons:De_minimis
[wiki-fair-use]: https://en.wikipedia.org/wiki/Wikipedia:FAQ/Copyright#What_is_fair_use?

@@

For a software license to be considered free/open,
it must be approved by either the [Open Source Initiative][osi-list] (OSI)
or the [Free Software Foundation][fsf-list] (FSF).
Some of the most common free/open software licenses are listed below:

@@

  [Link to full text][cc0-text]
  </details>
- <details>
  <summary>Creative Commons Attribution 4.0 (`CC-BY-4.0`)</summary>
  __Major conditions when sharing works__:

@@

  [Link to full text][cc-by-text]
  </details>
- <details>
  <summary>Creative Commons Attribution-ShareAlike 4.0 (`CC-BY-SA-4.0`)</summary>
  __Major conditions when sharing works__:

@@

  [Link to full text][cc-by-sa-text]
  </details>
::::

[cc0-text]: https://creativecommons.org/publicdomain/zero/1.0/legalcode#legal-code-title
[cc-by-text]: https://creativecommons.org/licenses/by/4.0/legalcode#legal-code-title
[cc-by-sa-text]: https://creativecommons.org/licenses/by-sa/4.0/legalcode#legal-code-title

Of course, if you encounter a Creative Commons license with a previous version
that satisfies the description above, it can be also considered free/open.
If you encounter a license not from Creative Commons and want to determine whether it is free/open,
you can consult the [lists of licenses accepted at Wikimedia Commons][commons-tags].
Note, however, that some of the licenses there might be inappropriate for text and media.
For example, licenses for software or for documentation commonly require
to distribute a work together with the full license text,
which is impractical for usage in print media.

[commons-tags]: https://commons.wikimedia.org/wiki/Commons:Copyright_tags

@@

  [Link to full text][cc-by-sa-text]
  </details>
::::

@@

If you encounter a license not listed here and want to determine whether it is free/open,
you can consult the [lists of licenses accepted at Wikimedia Commons][commons-tags].
Note, however, that some of the licenses there might be inappropriate for data,
because they do not mention _sui generis_ database rights.
:::

@@

You may have noticed that in many cases a version number is part of the license name.
This is because the organizations that created the licenses sometimes publish updated versions
to accommodate for developments in copyright law and the communities that use the licenses.
For example, the Creative Commons licenses (that start with `CC`) were [first published in 2002][cc-v1].
Since then, the possibility to relicense under later ([v2.0][cc-v2], 2004)
and compatible licenses ([v3.0][cc-v3], 2007) has been added,
a 30-day window to correct license violations has been established to combat [copyleft trolls][commons-trolls],
and _sui generis_ database rights have been covered explicitly ([v4.0][cc-v4], 2013).
There are many more [subtle differences between license versions][cc-versions],
therefore it is important to indicate which license version exactly one is referring to,
as the license of a work does not "update" automatically
(i.e., you have to follow the terms of that exact version).
CC0, officially introduced in 2009, is still at version [1.0][cc0-v1].

[cc-v1]: https://creativecommons.org/weblog/entry/3476
[cc-v2]: https://creativecommons.org/weblog/entry/4216#:~:text=Share%20Alike%20Across%20Borders
[cc-v3]: https://creativecommons.org/weblog/entry/7249#:~:text=BY-SA%20%E2%80%94%20Compatibility%20Structure%20Now%20Included
[cc-v4]: https://creativecommons.org/weblog/entry/40768
[cc0-v1]: http://www.creativecommons.org/weblog/entry/13304
[cc-versions]: https://wiki.creativecommons.org/wiki/License_Versions
[commons-trolls]: https://commons.wikimedia.org/wiki/Commons:Copyleft_trolling

@@

Finally, if you verified that the work has been made available to you under a free/open license,
you need to understand the conditions under which you are allowed to include the work in your project folder
and share it with others -- regardless of whether you modified it.
In the following, we will explain the most important conditions of the licenses listed above.
For other licenses we recommend reading their text in full and/or consulting with a lawyer.
The conditions of many popular free/open licenses are also [summarized on ChooseALicense.com][cal-appendix],
which you can use to get a general idea of a license
(but which does not obviate reading the license text in full).
Specifically for software licenses the _Open Source Automation Development Lab_
provides [useful checklists detailing the respective obligations][osadl].[^osadl-note]

[cal-appendix]: https://choosealicense.com/appendix/
[osadl]: https://www.osadl.org/Checklists/
[^osadl-note]: Accessing the checklists requires a (free) account on their website.

@@

Licenses may require that you provide attribution to the author(s)[^others]
and to retain any copyright notices,[^example-notice]
although the specifics vary between licenses.
For example, Creative Commons licenses other than CC0 require
that the name of the author, the title of the work, and a link to the work are provided.

[^others]: or others whose names are provided with the original work
[^example-notice]: Such as `Copyright (c) 2025 Jane Doe`

@@

If you modified the work and put a sufficient amount of creativity into it,
the result is called a _derivative work_.
It is (again) protected by copyright and even if you share it,
nobody except you is allowed to copy, modify, or distribute it further.
You can put it under a free/open license but don't have to.[^still-abide]
The only exception is if the original work has been made available under a "copyleft" license.
In that case, if you share the work with others
you must put it under the same license as the original work[^or-compatible]
-- or you commit a copyright violation.[^not-automatic]
This explains the origin of the term _copyleft_,
as it "turns copyright around" to make works permanently free: ©\ →\ 🄯.
In the realm of Creative Commons licenses, copyleft is also called _share alike_.
The opposite of a copyleft license is sometimes also called a _permissive_ license.

[^still-abide]: Of course, you still need to abide
by the license conditions of the original work, such as attribution.

[^or-compatible]: or sometimes also under a later version or under a compatible license

[^not-automatic]: Note, however, that [copyleft licenses are not automatic][commons-automatic]:
A derivative work of a copylefted work is not "implicitly" under the same license
-- others would commit a copyright violation if they just took it.
Only if the derivative author explicitly makes it available under the same license,
others may use the derivative work under the same terms.

[commons-automatic]: https://commons.wikimedia.org/wiki/Commons:Viral_licenses_are_not_automatic

What is considered creation of a derivative work varies between jurisdictions,
but the following are typical examples:
translating a text into a different language,
taking a picture of an artwork,
adding a song to a video, adapting a computer program to fit own use cases,
or taking a [screenshot of a computer program][commons-screenshots].
In all of these cases, if the original work is under a copyleft license,
then sharing of the resulting derivative work is only allowed
if it is put under the same license.

[commons-screenshots]: https://commons.wikimedia.org/wiki/Commons:Screenshots

@@

Obviously, if you are in doubt whether a condition of the license applies in your use case,
it is better to either follow it or to ask the author for clarification or explicit permission.
This is especially important given the existence of [_copyleft trolls_][commons-trolls]:
People who distribute works under a free/open license
but send out statutory damages claims after minor violations
without providing reusers an opportunity to fix their error.
In addition, in many European countries attribution may be necessary regardless
due to the author's moral rights.
For these reasons, we recommend that you provide attribution
for all works under a free/open license
-- even if the license does not strictly require it (e.g., CC0).
For this reason, you need to make note of every foreign work
included in your project folder and record its license.

@@

The free/open licenses we discuss here mostly grant copyrights
(with some of them explicitly granting patent rights and _sui generis_ database rights).
Therefore, you may lack other rights such as personality, privacy, moral, or trademark rights.
For example, [sharing photos that depict people][commons-photographs]
is not only a matter of copyright, but also of privacy rights.
Also, their commercial use may require the consent of the depicted person.
Finally, note that even if attribution is not a requirement of the license,
good scientific practice demands that appropriate citations are made.
:::

[commons-photographs]: https://commons.wikimedia.org/wiki/Commons:Photographs_of_identifiable_people

@@

1. The picture has been put under the license [CC0\ 1.0][cc0]
2. The license is free/open
3. The license is appropriate for images
4. The license contains no (major) conditions
:::

[cc0]: https://creativecommons.org/publicdomain/zero/1.0/
